Binning Strategies and Related Loss for Large Data

### 2015-04-30 Paper.Rnw
into the baseball example

\ktm{For this data example we will revisit the baseball data used earlier in this paper. To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning. Since the data are counts, each variable is recorded to an integer resolution so integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes. The data has only only 42583 observations, so we can bin finely without taking an inordinate amount of computation time. We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are 2 games X 10 strikeouts. The origin will be set to a half unit below the minimum value in each dimension, at ($0.5$,$-0.5$). }

\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot that with a continuous shading of the raw bin densities. We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts. In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities. Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale. Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure. }

>>>>

For this data example we will revisit the baseball data used earlier. To construct the binned scatterplot to display the relationship between strikeouts and game counts, we begin by specifying the bin dimensions and origin for standard rectangular binning. The data are counts, each variable is therefore recorded as an integer, corresponding to a resolution of one. Integer bin widths in each dimension of spatial bins will be used to avoid artificial density stripes. The data has only 42,583 observations, so we can bin finely without taking an inordinate amount of computation time. We elect to use approximately 50 bins in each dimension; the data range from 1 to 106 games and 0 to 513 strike-outs, thus we use bins that are two games wide and ten strikeouts high. The origin will be set to a half unit below the minimum value in each dimension, at ($0.5$,$-0.5$).

\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies. \hh{ XXX the legend needs fixing.  the numbers are running into one another. XXX be careful with `density' of the bins - the bins show frequencies not densities.} We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts. In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities. Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale. Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure. }
